\huge FOVEROS Chess Engine\\\Large (Modelling 2)

### 2014-11-28 Draft Final 1.tex
My latest

\subsubsection{Minimax Algorithm}
\indent\indent The generation of board states must be aided by an algorithm to evaluate and save the best board states. This is an adversarial search problem where only 2 players are competing. In order to make the optimal decision for a certain player, the MiniMax strategy is used. The idea is that in a certain player’s perspective, the objective is to maximise the player’s advantage and to minimise the opponent’s advantage. The generation starts from the “Initial State” and branches out until it reaches the “Leaf Nodes” or “Terminal Nodes” where the board is evaluated and the results backed up the tree as the recursion unwinds.\cite{chess1}
\\*\\
\indent Generation of board states however is a computationally expensive task especially to deeper depths. The branching factor can be defined as the number of children per node. The average branching factor for chess positions is 35 to 38 moves per position.  Therefore, 382 (1,444) game states need to be evaluated at depth 2, 383 (54,872) game states need to be evaluated at depth 3, 384 (2,085,136) game states need to be evaluated at depth 4. It is clear that the number of game states will continue to grow at an exponential rate.\cite{chess2}

@@

\indent The search then proceeds and references back to the alpha and beta value before it. If the values are found to be lower than the maximum lower bound or higher than the minimum upper bound, the values of alpha and beta are refreshed to the new values. However, pruning occurs when alpha is greater than beta and the branches leading from the selected node are not evaluated because it will not change the overall result of the tree. It will not change it because there is no overlap between the maximum lower bound and the minimum upper bound as shown in the figure.\cite{chess3}
\\*\\
\indent By implementing Alpha Beta pruning with random move ordering, the average number of nodes to be evaluated will be dropped to $b^{3d/4}$ from $b^d$. This means that Alpha Beta pruning is able to perform a depth 4 calculation at roughly the same speed as a depth 3 without Alpha Beta pruning.\cite{chess4}

@@

\begin{table}[h]
\centering
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
\multirow{2}{*}{Depth} & \multicolumn{2}{l|}{\begin{tabular}[c]{@{}l@{}}Without Alpha\\   Beta Pruning\end{tabular}} & \multicolumn{2}{l|}{\begin{tabular}[c]{@{}l@{}}With Alpha\\   Beta Pruning\end{tabular}} & \multicolumn{2}{l|}{\begin{tabular}[c]{@{}l@{}}Percentage Reduction Between\\   With and Without Pruning\end{tabular}} \\ \cline{2-7} 
                       & Time/s                                      & Nodes                               & Time/s                                 & Nodes                                 & Time                                 & Node                                \\ \hline
1                      & 0.046                                       & 20                                            & 0.05                                   & 20                                              & 8.695652174                          & 0                                   \\ \hline
2                      & 1.45                                        & 620                                           & 0.31                                   & 136                                             & -78.62068966                         & -78.06451613                        \\ \hline
3                      & 46.7                                        & 13928                                         & 1.98                                   & 832                                             & -95.76017131                         & -94.0264216                         \\ \hline
4                      & 1365                                        & 420180                                        & 23.4                                   & 8535                                            & -98.28571429                         & -97.96872769                        \\ \hline
5                      & \textgreater2400                            & \textgreater720000                            & 96                                     & 33022                                           &                                      &                                     \\ \hline
\end{tabular}
\caption{Performance results of $\alpha$-$\beta$ pruning}
\label{fig:chess1}
\end{table}

@@

\indent\indent The FOVEROS Engine which means “Awesome” in Greek employs basic understanding of the tree search algorithm, Alpha Beta pruning optimisation. It has also met the design aim of being a functional chessboard that has inbuilt rules and able to execute all the special moves without problems.

>>>>

\subsubsection{Minimax Algorithm}
\indent\indent The generation of board states must be aided by an algorithm to evaluate and save the best board states. This is an adversarial search problem where only 2 players are competing. In order to make the optimal decision for a certain player, the MiniMax strategy is used. The idea is that from a certain player’s perspective, the objective is to maximise the player’s advantage and to minimise the opponent’s advantage. The generation starts from the “Initial State” and branches out until it reaches the “Leaf Nodes” or “Terminal Nodes” where the board is evaluated and the results backed up the tree as the recursion unwinds.\cite{chess1}
\\*\\
\indent Generation of board states however is a computationally expensive task especially to deeper depths. The branching factor can be defined as the number of children per node. The average branching factor for chess positions is 35 to 38 moves per position.  Therefore, 38\textsuperscript{2} (1,444) game states need to be evaluated at depth 2, 38\textsuperscript{3} (54,872) game states need to be evaluated at depth 3, 38\textsuperscript{4} (2,085,136) game states need to be evaluated at depth 4. It is clear that the number of game states will continue to grow at an exponential rate.\cite{chess2}

@@

\indent The search then proceeds and references back to the alpha and beta value before it. If the values are found to be lower than the maximum lower bound or higher than the minimum upper bound, the values of alpha and beta are reassigned to the new values. However, pruning occurs when alpha is greater than beta and the branches leading from the selected node are not evaluated because it will not change the overall result of the tree. It will not change it because there is no overlap between the maximum lower bound and the minimum upper bound as shown in the figure.\cite{chess3}
\\*\\
\indent By implementing Alpha Beta pruning with random move ordering, the average number of nodes to be evaluated will be dropped to $b^{3d/4}$ from $b^d$. This means that Alpha Beta pruning is able to perform a depth 4 calculation at roughly the same speed as a depth 3 without Alpha Beta pruning.\cite{chess4}

@@

\begin{table}[h]
\centering
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
\multirow{2}{*}{Depth} & \multicolumn{2}{l|}{\begin{tabular}[c]{@{}l@{}}Without Alpha\\   Beta Pruning\end{tabular}} & \multicolumn{2}{l|}{\begin{tabular}[c]{@{}l@{}}With Alpha\\   Beta Pruning\end{tabular}} & \multicolumn{2}{l|}{\begin{tabular}[c]{@{}l@{}}Percentage Reduction Between\\   With and Without Pruning\end{tabular}} \\ \cline{2-7} 
                       & Time/s                                      & Nodes                               & Time/s                                 & Nodes                                 & Time                                 & Node                                \\ \hline
1                      & 0.046                                       & 20                                            & 0.05                                   & 20                                              & 8.70                          & 0                                   \\ \hline
2                      & 1.45                                        & 620                                           & 0.31                                   & 136                                             & -78.6                         & -78.1                        \\ \hline
3                      & 46.7                                        & 13928                                         & 1.98                                   & 832                                             & -95.8                         & -94.0                         \\ \hline
4                      & 1365                                        & 420180                                        & 23.4                                   & 8535                                            & -98.3                        & -98.0                        \\ \hline
5                      & \textgreater2400                            & \textgreater720000                            & 96                                     & 33022                                           &  -                                    &     -                                \\ \hline
\end{tabular}
\caption{Performance results of $\alpha$-$\beta$ pruning}
\label{fig:chess1}
\end{table}

@@

\indent\indent The FOVEROS Engine which means “Awesome” in Greek employs basic understanding of the tree search algorithm, MiniMax Theory, Alpha Beta pruning optimisation. It has also met the design aim of being a functional chessboard that has inbuilt rules and able to execute all the special moves without problems.
